Schema Evolution in Interactive Programming Systems

### 2024-10-25 paper.tex
More final version
Make conference publishing happier

\usepackage{csquotes}
\usepackage{changepage}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{ccicons}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{framed}

\lstdefinelanguage[programming]{TeX}[AlLaTeX]{TeX}{%
  deletetexcs={title,author,bibliography},%
  deletekeywords={tabular},
  morekeywords={abstract},%
  moretexcs={chapter},%
  moretexcs=[2]{title,author,subtitle,keywords,maketitle,titlerunning,authorinfo,affiliation,authorrunning,paperdetails,acks,email},
  moretexcs=[3]{addbibresource,printbibliography,bibliography},%
}%
\lstset{%
  language={[programming]TeX},%
  keywordstyle=\firamedium,
  belowskip=1em,
  aboveskip=1em,
  framesep=0.5em,
  basicstyle=\normalsize\ttfamily,
  %stringstyle=\color{RosyBrown},%
  %texcsstyle=*{\color{Purple}\mdseries},%
  %texcsstyle=*[2]{\color{Blue1}},%
  %texcsstyle=*[3]{\color{ForestGreen}},%
  %commentstyle={\color{FireBrick}},%
  numbersep=1em,
  xleftmargin=1.5em,
  escapechar=`,}
\newcommand*{\CTAN}[1]{\href{http://ctan.org/tex-archive/#1}{\nolinkurl{CTAN:#1}}}
%%

@@

\DeclareRobustCommand{\frameworkbox}[1]{\leftbar#1\endleftbar}
\newcommand{\frameworkboxtitle}[1]{\noindent{\firamedium #1.}\quad}

@@

@@

\author[e]{Geoffrey Litt}[0000-0003-0858-5165]
\authorinfo{is a senior researcher at the independent research lab Ink \& Switch.
Previously, he completed a PhD at MIT CSAIL in the Software Design Group advised by Daniel Jackson.
Before that, he spent five years as an early engineer and designer building the edtech startup Panorama Education (YC S13).
Contact him at \email{gklitt@gmail.com}.}
\affiliation[e]{Ink \& Switch, Planet Earth}

@@

\vspace{-0.4em}
\paragraph{Locality of Change.}
We can view diverse real-world schema evolution problems from a common perspective by distinguishing between two different dimensions of schema change.
The \emph{local dimension} is concerned with how schema evolution affects data and code within
a single program. The \emph{non-local dimension} is concerned with how schema evolution
propagates across different program variants that may exist independently.

\vspace{-0.4em}
\paragraph{Program Layers.}
We consider the more permanent layer of \emph{schema} and two less permanent layers of code and
data that are structured by those schema. In specific case studies, the layers may map to
different aspects of programs, some may be not present, and some may be implicit.

@@

\vspace{-0.4em}
\paragraph{Program Variants.}
In programming systems that enable collaboration, multiple variants of a program may exist
concurrently and their schema, data and code may need to be synchronized. Although collaboration
is missing from some of our case studies, it adds an important dimension that is crucial for
schema evolution in future programming systems. We consider two characteristics.

@@

\vspace{-0.4em}
\paragraph{Challenge Problems.}
These dimensions, layers, and variants generate a wide range of problem configuration.
Our aim is not to cover all of them. Instead we use this perspective to recast a number of
existing challenges in real-world programming systems as instances of the same problem of
schema evolution. The case studies show how well-known solutions or problems in one domain
map to another domain. We hope they provide an inspiration and a benchmark for
developers of future systems.

@@

\begin{table}[t]
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Add & case to \texttt{Event} & Requires adding corresponding case to \texttt{update} \\
Remove & case from \texttt{Event} & Remove unused code from \texttt{update} \\
Add & field to \texttt{State} & Migrate state value and initialize field \\
Remove & field from \texttt{State} & Migrate state (assuming field unused) \\
Modify & structure of \texttt{State} & Migrate state value and edit code accordingly \\
\bottomrule
\end{tabular}
\vspace{0.3em}
\caption{Changes to types and how a programming system should handle them}
\label{tbl:elmchanges}
\vspace{-1em}
\end{table}

@@

\begin{table}[t]
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Extract & new type of entity & Migrate data to a new table and add links \\
Absorb & a linked type of entity & Migrate data and remove unneeded table \\
Merge & multiple data items & Automatic de-duplication with mistake handling \\
Split & multiple data items & Interactively, following user guidance\\
\bottomrule
\end{tabular}
\vspace{0.3em}
\caption{Changes to types and how a programming system should handle them}
\label{tbl:dbchanges}
\vspace{-1em}
\end{table}

@@

\begin{table}[t]
\begin{adjustwidth}{-2em}{-2em}
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Change & Tag of an element & Apply to all matching elements and affected formulas\\
Wrap & Selected element(s) & Wrap elements and update selectors in affected formulas  \\
Split & Selected element(s) & Split matching elements using a rule (regular expression)\\
Add & New item to a list & Add data without affecting document structure (schema)\\
Reorder & List items & Reorder data without affecting document structure (schema)\\
\bottomrule
\end{tabular}
\vspace{0.3em}
\caption{Changes to schema and data and how the programming system should respond}
\label{tbl:docchanges}
\end{adjustwidth}
\vspace{-1em}
\end{table}

@@

\begin{table}[b]
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Add/Rename & state & Simple  \\
Remove & state & Structurally simple, but heuristic if state is current \\
Add & variable & Add entry to map and initialize \\
Remove & variable & Remove from the variable map (assuming it is unused)\\
Rename & variable & Rename in class, preserving value\\
Type change & variable & Migrate value in map \\
Add & transition & Simple \\
Remove & transition & Structurally simple, but heuristic for pending events\\
Event change & transition & Structurally simple, but heuristic for pending events\\
Add/remove & statement & Hotswap code at quiescent point of interpreter\\
\bottomrule
\end{tabular}
\vspace{0.3em}
\caption{Possible program changes and how to deal with them at run time}
\label{tbl:smchanges}
\vspace{-1em}
\end{table}

@@

\paragraph{Schema evolution.}
Schema evolution has long been a major problem for SQL databases because
the SQL Data Definition Language (DDL) has limited capabilities to redefine existing data. It can rename tables and columns. SQLite \cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
MYSql \cite{mysqlAlterTable} can reorder columns without destroying their data. But apart from such special cases, SQL databases have no general purpose support for data migration. As a result in practice schema evolution is mostly done by writing custom Data Manipulation Language (DML) code to migrate the data. Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
In \emph{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.

@@

\paragraph{Type-driven transformations.}
Programming language theory undergirds research on type-driven transformations. Bidirectional Transformations~\cite{czarnecki2009bidirectional} have been used for data format conversion and view update.
Coupled Transformations~\cite{Berdaguer07, alcino06, Cleve2006} express transformations coordinated between types and their instances by encoding them into functions on Haskell GADTs built with strategy combinators. \citet{JVisser08} extends the encoding to transform queries written in point-free style. \citet{lammel16} recapitulates coupled transformations within logic programming which extends it to transform logic programs. It is not clear how bidirectional transformations can handle generating the unique IDs required in some evolutions, nor the information loss and duplication which can arise in some divergence scenarios.

\paragraph{Model Driven Engineering.}
Model Driven Engineering must also handle evolution. Unlike the textual artifacts involved in other domains, models typically assign unique identifiers enabling more precise differencing and mergeing~\cite{alanen2003}.
Models are often themselves modeled by a metamodel, which lifts the evolution problem to the metalevel: models must be migrated through the evolution of their metamodel, analogously to changing the grammar of a programming language.~\cite{Herrmannsdoerfer11}. \citet{Cicchetti11} study metamodel evolution in the context of divergent changes.
\citet{vermolen11} define a DSL for evolving a data model by generating SQL migrations on the backing database. They consider the evolution of OO-style subclass hierarchies which goes beyond most other work. Their running example includes both our \emph{Extract Entity} and \emph{Multiplicity Change} examples, and would make for a good follow-on problem. However they do not address code/query rewriting.

\paragraph{Migration by example.}
Programming by example has been used to infer schema migration programs from examples specified by the developer~\cite{wang20, Alexe11}.
This work deals with the reality that schema live in the database and so examples must be supplied separately, but we wonder why not incorporate examples into a schema definition language? Textual edits to schema in this language would also adapt the examples, which then could be used to infer an abstract migration, thus avoiding the ambiguities of textually comparing pure schema.

\paragraph{Data wrangling.} The evolution of data formats is a concern in Data Science. Data~Diff~\cite{sutton18} compares two versions of tabular data and finds a minimal matching transformation.
Wrangler~\cite{kandel11} observes interactive manipulation of sampled data to write data transformation programs.
\citet{petersohn20} present an algebra of operations on Python dataframes that include shape transformations.

\paragraph{Live programming.}
Live programming~\cite{tanimoto90, rein2018exploratory, HazelnutLive19} seeks to speed up feedback in (primarily solo) programming.
% \cite{mcdirmid13, Rauch_2019, beckmann2021shortening}
\emph{Hot reloading/swapping}~\cite{barenz2020essence, hicks2005dynamic} installs code changes into a running system preserving some runtime state but not attempting to migrate that state through type changes.
\citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.
\citet{RuntimeConstraint} study the live modeling problem in Section~\ref{sec:livemodeling}. They use constraints supplied by the developer to solve for the best new runtime state following a DSL language evolution. \citet{vanRozen19} take an alternative approach to the same problem using \emph{origin tracking} in textual editing.
%\citet{vanRozen23} subsequently proposed a metaprogramming language for live DSL programming.

>>>>

\usepackage{csquotes}
\usepackage{changepage}
\usepackage{multicol}
\usepackage{xcolor}
%\usepackage{ccicons}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{framed}

\lstdefinelanguage[programming]{TeX}[AlLaTeX]{TeX}{%
  deletetexcs={title,author,bibliography},%
  deletekeywords={tabular},
  morekeywords={abstract},%
  moretexcs={chapter},%
  moretexcs=[2]{title,author,subtitle,keywords,maketitle,titlerunning,authorinfo,affiliation,authorrunning,paperdetails,acks,email},
  moretexcs=[3]{addbibresource,printbibliography,bibliography},%
}%
\lstset{%
  language={[programming]TeX},%
  keywordstyle=\firamedium,
  belowskip=1em,
  aboveskip=1em,
  %framesep=0.5em,
  basicstyle=\normalsize\ttfamily,
  %stringstyle=\color{RosyBrown},%
  %texcsstyle=*{\color{Purple}\mdseries},%
  %texcsstyle=*[2]{\color{Blue1}},%
  %texcsstyle=*[3]{\color{ForestGreen}},%
  %commentstyle={\color{FireBrick}},%
  numbersep=1em,
  %xleftmargin=1.5em,
  escapechar=`,}
\newcommand*{\CTAN}[1]{\href{http://ctan.org/tex-archive/#1}{\nolinkurl{CTAN:#1}}}
%%

@@

\DeclareRobustCommand{\frameworkbox}[1]{\leftbar#1\endleftbar}
\newcommand{\frameworkboxtitle}[1]{\noindent{\firamedium #1}\quad}

@@

\paperdetails{
  submitted=2024-05-31,
  published=2024-10-15,
  year=2025,
  volume=9,
  issue=1,
  articlenumber=2,
}

@@

\author[e]{Geoffrey Litt}[0000-0003-0858-5165]
\authorinfo{is a senior researcher at the independent research lab Ink \& Switch.
Previously, he completed a PhD at MIT CSAIL in the Software Design Group advised by Daniel Jackson.
Before that, he spent five years as an early engineer and designer building the edtech startup Panorama Education (YC S13).
Contact him at \email{gklitt@gmail.com}.}
\affiliation[e]{Ink \& Switch, USA}

@@

\vspace{-0.4em}
\paragraph{Locality of Change}
We can view diverse real-world schema evolution problems from a common perspective by distinguishing between two different dimensions of schema change.
The \emph{local dimension} is concerned with how schema evolution affects data and code within
a single program. The \emph{non-local dimension} is concerned with how schema evolution
propagates across different program variants that may exist independently.

\vspace{-0.4em}
\paragraph{Program Layers}
We consider the more permanent layer of \emph{schema} and two less permanent layers of code and
data that are structured by those schema. In specific case studies, the layers may map to
different aspects of programs, some may be not present, and some may be implicit.

@@

\vspace{-0.4em}
\paragraph{Program Variants}
In programming systems that enable collaboration, multiple variants of a program may exist
concurrently and their schema, data and code may need to be synchronized. Although collaboration
is missing from some of our case studies, it adds an important dimension that is crucial for
schema evolution in future programming systems. We consider two characteristics.

@@

\vspace{-0.4em}
\paragraph{Challenge Problems}
These dimensions, layers, and variants generate a wide range of problem configuration.
Our aim is not to cover all of them. Instead we use this perspective to recast a number of
existing challenges in real-world programming systems as instances of the same problem of
schema evolution. The case studies show how well-known solutions or problems in one domain
map to another domain. We hope they provide an inspiration and a benchmark for
developers of future systems.

@@

\begin{table}[t]
\caption{Changes to types and how a programming system should handle them}
\label{tbl:elmchanges}
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Add & case to \texttt{Event} & Requires adding corresponding case to \texttt{update} \\
Remove & case from \texttt{Event} & Remove unused code from \texttt{update} \\
Add & field to \texttt{State} & Migrate state value and initialize field \\
Remove & field from \texttt{State} & Migrate state (assuming field unused) \\
Modify & structure of \texttt{State} & Migrate state value and edit code accordingly \\
\bottomrule
\end{tabular}
\vspace{-0.5em}
\end{table}

@@

\begin{table}[t]
\caption{Changes to schema and how a programming system should handle them}
\label{tbl:dbchanges}
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Extract & new type of entity & Migrate data to a new table and add links \\
Absorb & a linked type of entity & Migrate data and remove unneeded table \\
Merge & multiple data items & Automatic de-duplication with mistake handling \\
Split & multiple data items & Interactively, following user guidance\\
\bottomrule
\end{tabular}
\vspace{-0.5em}
\end{table}

@@

\begin{table}[t]
\caption{Changes to schema and data and how the programming system should respond}
\label{tbl:docchanges}
\begin{adjustwidth}{-2em}{-2em}
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Change & Tag of an element & Apply to all matching elements and affected formulas\\
Wrap & Selected element(s) & Wrap elements and update selectors in affected formulas  \\
Split & Selected element(s) & Split matching elements using a rule (regular expression)\\
Add & New item to a list & Add data without affecting document structure (schema)\\
Reorder & List items & Reorder data without affecting document structure (schema)\\
\bottomrule
\end{tabular}
\end{adjustwidth}
\vspace{-0.5em}
\end{table}

@@

\begin{table}[b]
\caption{Possible program changes and how to deal with them at run time}
\label{tbl:smchanges}
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Add/Rename & state & Simple  \\
Remove & state & Structurally simple, but heuristic if state is current \\
Add & variable & Add entry to map and initialize \\
Remove & variable & Remove from the variable map (assuming it is unused)\\
Rename & variable & Rename in class, preserving value\\
Type change & variable & Migrate value in map \\
Add & transition & Simple \\
Remove & transition & Structurally simple, but heuristic for pending events\\
Event change & transition & Structurally simple, but heuristic for pending events\\
Add/remove & statement & Hotswap code at quiescent point of interpreter\\
\bottomrule
\end{tabular}
\vspace{-0.5em}
\end{table}

@@

\paragraph{Schema Evolution}
Schema evolution has long been a major problem for SQL databases because
the SQL Data Definition Language (DDL) has limited capabilities to redefine existing data. It can rename tables and columns. SQLite \cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
MYSql \cite{mysqlAlterTable} can reorder columns without destroying their data. But apart from such special cases, SQL databases have no general purpose support for data migration. As a result in practice schema evolution is mostly done by writing custom Data Manipulation Language (DML) code to migrate the data. Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
In \emph{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.

@@

\paragraph{Type-Driven Transformations}
Programming language theory undergirds research on type-driven transformations. Bidirectional Transformations~\cite{czarnecki2009bidirectional} have been used for data format conversion and view update.
Coupled Transformations~\cite{Berdaguer07, alcino06, Cleve2006} express transformations coordinated between types and their instances by encoding them into functions on Haskell GADTs built with strategy combinators. \citet{JVisser08} extends the encoding to transform queries written in point-free style. \citet{lammel16} recapitulates coupled transformations within logic programming which extends it to transform logic programs. It is not clear how bidirectional transformations can handle generating the unique IDs required in some evolutions, nor the information loss and duplication which can arise in some divergence scenarios.

\paragraph{Model Driven Engineering}
Model Driven Engineering must also handle evolution. Unlike the textual artifacts involved in other domains, models typically assign unique identifiers enabling more precise differencing and mergeing~\cite{alanen2003}.
Models are often themselves modeled by a metamodel, which lifts the evolution problem to the metalevel: models must be migrated through the evolution of their metamodel, analogously to changing the grammar of a programming language.~\cite{Herrmannsdoerfer11}. \citet{Cicchetti11} study metamodel evolution in the context of divergent changes.
\citet{vermolen11} define a DSL for evolving a data model by generating SQL migrations on the backing database. They consider the evolution of OO-style subclass hierarchies which goes beyond most other work. Their running example includes both our \emph{Extract Entity} and \emph{Multiplicity Change} examples, and would make for a good follow-on problem. However they do not address code/query rewriting.

\paragraph{Migration by Example}
Programming by example has been used to infer schema migration programs from examples specified by the developer~\cite{wang20, Alexe11}.
This work deals with the reality that schema live in the database and so examples must be supplied separately, but we wonder why not incorporate examples into a schema definition language? Textual edits to schema in this language would also adapt the examples, which then could be used to infer an abstract migration, thus avoiding the ambiguities of textually comparing pure schema.

\paragraph{Data Wrangling} The evolution of data formats is a concern in Data Science. Data~Diff~\cite{sutton18} compares two versions of tabular data and finds a minimal matching transformation.
Wrangler~\cite{kandel11} observes interactive manipulation of sampled data to write data transformation programs.
\citet{petersohn20} present an algebra of operations on Python dataframes that include shape transformations.

\paragraph{Live Programming}
Live programming~\cite{tanimoto90, rein2018exploratory, HazelnutLive19} seeks to speed up feedback in (primarily solo) programming.
% \cite{mcdirmid13, Rauch_2019, beckmann2021shortening}
\emph{Hot reloading/swapping}~\cite{barenz2020essence, hicks2005dynamic} installs code changes into a running system preserving some runtime state but not attempting to migrate that state through type changes.
\citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.
\citet{RuntimeConstraint} study the live modeling problem in Section~\ref{sec:livemodeling}. They use constraints supplied by the developer to solve for the best new runtime state following a DSL language evolution. \citet{vanRozen19} take an alternative approach to the same problem using \emph{origin tracking} in textual editing.
%\citet{vanRozen23} subsequently proposed a metaprogramming language for live DSL programming.
